### 2022-02-28 paper.Rmd
continued

In the following, I present a variation of the $\mathit{SSE}$ that quantifies how the variance of numeric data differs between groups. To motivate the new criterion, we first consider a set of one-dimensional observations $(x_1, \ldots, x_N)$. The extension to the general case of $P$ attributes will be straight-forward.

The sample variance of a set of observations $x = (x_1, \ldots, x_N)$ is computed as the mean of the squared distances between observations and their mean $\overline{\mathbf{x}}$:

@@

@@

It is unlikely that an exclusive optimization of $\mathit{SSE_{Var}}$ (i.e., equalizing variances) is of interest without also considering the $\mathit{SSE}$ (i.e., equalizing means). A combination of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$, however, can be employed to simultaneously optimize both objectives, which is a reasonable idea when striving for overall cluster similarity. The most simple approach for such a bicriterion optimization is the weighted sum approach [@naidu2014; @marler2010]. That is, both criteria are computed independently from each other and then combined into a single criterion through a weighted sum:

$$
\mathit{SSE_{kplus}} = w_1 
  \sum\limits_{j=1}^{P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (y_{ij} - \overline{x}_j ^{(k)})^2 + 
  w_2 \sum\limits_{j=1}^{P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (x_{ij} - \overline{y}_j ^{(k)})^2
$$

The weights $w_1$ and $w_2$ can be chosen to adjust the relative importance of equalizing means and variances, respectively. Some simplifications can be arranged when ignoring the weights $w_1$ and $w_2$, e.g., by setting them both to 1. When defining $z_i = (x_{i1}, \ldots, x_{iP}, y_{i1}, \ldots, x_{iP})$, i.e., the concatenation of the $x_{ij}$ and $y_{ij}$ values, the bicriterion $\mathit{SSE_{kplus}}$ is then computed as

@@

Note that a problem may arise if the weights $w_1$ and $w_2$ are ignored when optimizing the k-plus objective. Since the computation of the data matrix $\mathbf{Y}_{N \times P}$ involves squaring, the squared distances $y_{ij}$ have much larger values, on average, than the raw values $x_{ij}$ (also see Figure 1). Therefore, the $\mathit{SSE_{Var}}$ criterion naturally receives greater weight when computing $\mathit{SSE_{kplus}}$ than the standard $\mathit{SSE}$ criterion. This imbalance is most likely unwarranted because optimizing similarity with regard to variances is not more important than optimizing similarity with regard to means. A solution is to standardize all values $x_{ij}$ and $y_{ij}$ to the same scale before computing $SSE_{kplus}$, e.g. via $z$-standardization.

## Higher order moments, skewness and kurtosis

The outlined logic to equalize variance in different groups is easily generalized to higher order moments such as skewness and kurtosis, i.e., the third and fourth distribution moments, respectively. Generally, the *j*'th sample moment of a variable $x = (x_1, \ldots, x_N)$ is computed as [@joanes1998]

@@

where $j = 2$ gives the variance, $j = 3$ the skewness, and $j = 4$ the kurtosis.

The approach outlined in sections 1.1 and 1.2 illustrated how the definition of the variance can be used to add variables to the data set to minimize differences with regard to the variance between groups. Changing the power of the deviation between mean and data points, we can use the same logic to equalize any desired sample moment between groups. I expect this to be interesting in particular for the skewness and kurtosis to minimize differences with regard to distribution asymmetry (skewness) and the propensity to include outliers (kurtosis).

## Covariance structure

@@

Some clarifications with respect to nomenclature are needed. The k-plus criterion is not a single objective function for anticlustering, but instead is a familiy of objectives that use the standard k-means criterion $SSE$ in order to minimize differences between groups with regard to additional characterizations of a data set; that is, not just the mean, but also the variance, covariances, or higher order moments such as skewness and kurtosis. Which of these objectives are pursued depends on a researchers needs. In the most basic---and arguably most important---case, k-plus can be used to not just minimize differences with regard to the mean of variables but also their variance. Measures of location (mean) and spread (variance / standard deviation) are arguably most important to researchers when they "eyeball" their data and they are routinely reported as characterizations of a data set. Hence, by default, the k-plus anticlustering objective implies that means and variances should be equal between groups.

While less routinely reported (and less well understood), skewness are kurtosis are also important descriptive characterizations of a data set [@decarlo1997; @westfall2014]. To optimize overall similarity between groups, it may thus be desirable to also minimize differences with regard to these distribution moments. In principle, any higher order moments can also be included as part of the k-plus objective. Note, however, that the inclusion each additional moment requires appending as many features to the data set as there were to begin with.

>>>>

In the following, I present a variation of the $\mathit{SSE}$ that quantifies how the variance of numeric data differs between groups. To motivate the new criterion, we first consider a set of one-dimensional observations $(x_1, \ldots, x_N)$. The extension to the general case of $P$ attributes will be straight forward.

The sample variance[^biasedestimate] of a set of observations $x = (x_1, \ldots, x_N)$ is computed as the mean of the squared distances between observations and their mean $\overline{\mathbf{x}}$:

@@

[^biasedestimate]: this

@@

It is unlikely that an exclusive optimization of $\mathit{SSE_{Var}}$ (i.e., equalizing variances) is of interest without also considering the $\mathit{SSE}$ (i.e., equalizing means). A combination of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$, however, can be employed to simultaneously optimize both objectives, which is a reasonable idea when striving for overall between-group similarity. The most simple approach for such a bicriterion optimization is the weighted sum approach [@naidu2014; @marler2010]. That is, both criteria are computed independently from each other and then combined into a single criterion through a weighted sum:

$$
w_1 \sum\limits_{j=1}^{P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (x_{ij} - \overline{x}_j ^{(k)})^2 + 
w_2 \sum\limits_{j=1}^{P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (y_{ij} - \overline{y}_j ^{(k)})^2
$$

The weights $w_1$ and $w_2$ can be chosen to adjust the relative importance of equalizing means and variances, respectively. Some simplifications can be arranged when ignoring the weights $w_1$ and $w_2$, e.g., by setting them both to 1. When defining $z_i = (x_{i1}, \ldots, x_{iP}, y_{i1}, \ldots, x_{iP})$, i.e., the concatenation of the $x_{ij}$ and $y_{ij}$ values, the bicriterion $\mathit{SSE_{kplus}}$ can be computed as

@@

Note that a problem may arise if the weights $w_1$ and $w_2$ are ignored when optimizing the k-plus objective. Since the computation of the data matrix $\mathbf{Y}_{N \times P}$ involves squaring, the squared distances $y_{ij}$ have much larger values, on average, than the raw values $x_{ij}$. Therefore, the $\mathit{SSE_{Var}}$ criterion naturally receives greater weight when computing $\mathit{SSE_{kplus}}$ than the standard $\mathit{SSE}$ criterion. This imbalance is most likely unwarranted because optimizing similarity with regard to variances is not more important than optimizing similarity with regard to means. A solution is to standardize all values $x_{ij}$ and $y_{ij}$ to the same scale before computing $SSE_{kplus}$, e.g. via $z$-standardization.

## Skewness, kurtosis, and higher order moments

The outlined logic to equalize the variance, which is the second moment of a distribution, can be generalized to higher order moments in a straight forward manner. The *j*'th sample moment of a variable $x = (x_1, \ldots, x_N)$ can be computed as

@@

where $j = 2$ gives the variance, $j = 3$ the skewness, and $j = 4$ the kurtosis [@joanes1998]. Thus, by changing $j$---the power of the deviation between mean and data points---we can use the k-plus logic to equalize any desired sample moment between groups. For each moment $j$ and for each variable $x$, this is accomplished by adding a new variable $y^{(j)} = (y^{(j)}_1, \ldots, y^{(j)}_N)$ with $y^{(j)}_i = (x_i - \overline{\mathbf{x}})^j$ to the data set and subsequently applying standard k-means anticlustering. I expect this to be particularly interesting for the skewness and kurtosis when aiming to minimize differences with regard to distribution asymmetry (skewness) and the propensity to include outliers (kurtosis).

## Covariances

@@

Some clarifications with respect to nomenclature are needed. The k-plus criterion is not a single objective function for anticlustering, but instead a familiy of objectives that extend the standard k-means objective ($SSE$) to minimize differences between groups with regard to several characterizations of a data set. That is: not just the mean, but also the variance, covariances, or higher order moments such as skewness and kurtosis. Which of these objectives are pursued depends on a researchers needs. In the most basic---and arguably most important---case, k-plus anticlustering can be used to not just minimize differences with regard to the mean of variables but also their variance. Measures of location (mean) and spread (variance / standard deviation) are arguably most important to researchers when they "eyeball" their data and they are routinely reported as characterizations of a data set. Hence, by default, the k-plus anticlustering objective implies that means and variances should be equal between groups.

While less routinely reported, skewness are kurtosis are also important descriptive characterizations of a data set [@decarlo1997; @westfall2014]. To optimize overall similarity between groups, it may thus be desirable to also minimize differences with regard to these distribution moments. In principle, any higher order moments can also be included as part of the k-plus objective. Note, however, that the inclusion each additional moment requires appending as many features to the data set as there were to begin with.
